|

### 2024-09-05 manuscript/paperCMSACversion.Rmd
most updated.

Fighting in the National Hockey League (NHL) has been an integral part of the sport since the earliest days of the league. More than just a way to express frustration, players often use fighting strategically to the benefit of the team. One of these strategies includes instigating a fight when a player from the opposite team has violated an unwritten rule of the game, which historically has led to hockey teams carrying so-called "enforcers" on their rosters. The enforcer is the teammember tasked with fighting these opposing players. While this is not an officially defined position, the nature of needing to be able to fight any opponent generally designates the biggest and strongest players as those most likely to be enforcers.

@@

While any player on a hockey team can get into a fight, enforcers are exposed to many more fights than the typical NHL player. As a result of being exposed to additional contact, they are potentially exposed to additional head trauma beyond the amount associated with playing hockey. There are many questions about the long-term implications of this head trauma and its potential to result in negative health outcomes in the future. The ingrained nature of fighting into the game of hockey is greatly criticized and there are many active debates about its current declining trend. Figure \ref{fig:enforcer_density} shows a decline in the proportion of enforcers to non-enforcers as players get younger. The question is, should the game continue to move away from contact-driven strategy in favor of younger, faster, highly-skilled players?

```{r enforcer_density, echo = FALSE, fig.cap='Density Plot of Study Cohorts by DOB'}
cohort_dat = dat %>% mutate(cohort = case_when(PIM.GP > 3 ~ 'Enforcer',
                                  Total.Fights > 50 ~ 'Enforcer',
                                  .default = 'Non-Enforcer'))

cohort_dat$cohort = as.factor(cohort_dat$cohort)

ggplot(cohort_dat, aes(x = DOB)) + geom_density(aes(fill = cohort),alpha = 0.75 )
```

@@

As mentioned earlier, the outcome of this study is relevant to the debates about regulations and discourse around head trauma in hockey. An accurate analysis is imperative to prevent the spread of misinformed conclusions and best inform regulatory safety practices going forward.

... talk about our conclusions

@@

 \hline
\end{tabular}
\end{center}
\caption{Comparing enforcers to non-enforcers when enforcer is defined as E-F \label{tab:EDAEF}}
\end{table}

@@

\caption{Comparing enforcers to non-enforcers when enforcer is defined as E-P \label{tab:EDAEP}}
\end{table}

@@

# Results {#sec:results}
Table \ref{tab:EF} shows the results for the Cox proportional hazards model when enforcers were defined based on their number of career fights.  No statistically significant difference was observed in the hazard functions of enforcers versus non-enforcers (p-value = 0.549). However, a significant difference is observed between enforcers and non-enforcers when the enforcer is defined based on their number of penalty minutes per game (p-value = 0.00169).  The estimated hazard ratio for E-P is 2.29, indicating a 129\% increase in the hazard for enforcers vs non-enforcers when all other variables are held constant.

Figure \ref{fig:estsurv} shows the estimated survival curves for enforcers (E-P) versus non-enforcers for defenders with all continuous variables held constant at mean levels.  Using this, the estimated survival rate at the age of 50 for enforcers vs non-enforcers is 96.57\% (95\% CI: 94.63\%, 98.56\%) vs 98.49\% (95\% CI: 97.98\%, 99.00\%).  For the same conditions at the age of 75, the estimated survival rate for enforcers is about 74.4\% (95\% CI: 61.4\%, 90.32\%) compared to non-enforcers estimate survival rate of 87.95\% (95\% CI: 83.09\%, 93.10\%).  The estimated survival rate for enforcers drops an additional `r (96.57-74.4)-(98.49 - 87.95)`\% between the ages of 50 and 75 versus non-enforcers.

These results suggest that the conclusions of @Popkin2023 are overstated due to the differences in exposure time between the groups at the time of their final analysis.

join all from jama nhl data, condition on death from jama deaths and look at yob between enforcer/control
```{r}
cf = read.csv('../JAMA NHL Data/C-F-Table 1.csv')
cp = read.csv('../JAMA NHL Data/C-P-Table 1.csv')

ef = read.csv('../JAMA NHL Data/E-F-Table 1.csv')
ep = read.csv('../JAMA NHL Data/E-P-Table 1.csv')

jama_controls = dat %>% filter(Player %in% cf$Player | Player %in% cp$Player)

jama_enforcers = dat %>% filter(Player %in% ef$Player | Player %in% ep$Player)

jama_short = rbind(jama_controls, jama_enforcers)

jama_deaths = read.csv('../JAMA NHL Data/Deaths-Table 1.csv')

jama_short %>%
  filter(Player %in% jama_deaths$Enforcers) %>%
  mutate(enforcer = ifelse(EF == 1 | EP == 1, 1, 0)) %>% 
  group_by(enforcer) %>%
  summarize(mean(DOB.Year))

jama_short %>%
  mutate(deathage = as.numeric(difftime(Death.Date,
                                        Birth.Date,
                                        units = "weeks")/52.25)) %>%
  filter(Player %in% jama_deaths$Enforcers) %>%
  mutate(enforcer = ifelse(EF == 1 | EP == 1, 1, 0)) %>% 
  group_by(enforcer) %>%
  summarize(mean(deathage))

@@

```{r estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for E-P for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5)"}
coxphmod_EP <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

@@

We would like to thank @Popkin2023 for their generosity and willingness to share the data provided for the above analysis.

>>>>

Fighting in the National Hockey League (NHL) has been an integral part of the sport since the earliest days of the league. More than just a way to express frustration, players often use fighting strategically to the benefit of the team. One of these strategies includes instigating a fight when a player from the opposite team has violated an unwritten rule of the game, which historically has led to hockey teams carrying so-called "enforcers" on their rosters. An enforcer is a member of a hockey team who, along with being a regular player, is the one tasked with fighting these opposing players. While this is not an officially defined position, the nature of needing to be able to fight any opponent generally designates the biggest and strongest players as those most likely to be enforcers.

@@

While any player on a hockey team can get into a fight, enforcers are exposed to many more fights than the typical NHL player. As a result of being exposed to additional contact, they are potentially exposed to additional head trauma beyond the amount associated with playing hockey. There are many questions about the long-term implications of this head trauma and its potential to result in negative health outcomes in the future. The ingrained nature of fighting into the game of hockey is greatly criticized and there are many active debates about its current declining trend (can we cite something that shows the trend is declining?). Should the game continue to move away from fighting in favor of younger, faster, highly-skilled players? Objectively, however, fighting has historically been a non-negligible part of the game and correctly understanding the implications this has on player health may help direct the future of the sport.

@@

As mentioned earlier, the outcome of this study is relevant to the debates about regulations and discourse around head trauma in hockey. An accurate analysis is imperative to prevent the spread of misinformed conclusions. ... talk about our conclusions

@@

 \hline
\end{tabular}
\end{center}
\caption{Comparing enforcers to non enforcers when enforcer is defined as E-F \label{tab:EDAEF}}
\end{table}

@@

\caption{Comparing enforcers to non enforcers when enforcer is defined as E-P \label{tab:EDAEP}}
\end{table}

@@

# Results {#sec:results}
Table \ref{tab:EF} shows the results for the Cox proportional hazards model when enforcers were defined based on their number of career fights.  No statistically significant difference was observed in the hazard functions of enforcers versus non enforcers (p-value = 0.549). However, a significant difference is observed between enforcers and non-enforcers when the enforcer is defined based on their number of penalty minutes per game (p-value = 0.00169).  The estimated hazard ratio for E-P is 2.29, indicating a 129\% increase in the hazard for enforcers vs non-enforcers when all other variables are held constant.

Figure \@ref{fig:estsurv} shoes the estimated survival curves for enforcers versus non-enforcers (E-P) for defenders with all continuous variables held constant at mean levels.  Using this, the estimated survival rate at the age of 50 for enforcers vs non-enforcers is 96.57\% (95\% CI: 94.63\%, 98.56\%) vs 98.49\% (95\% CI: 97.98\%, 99.00\%).  For the same conditions at the age of 75, the estimated survival rate for enforcers is about 74.4\% (95\% CI: 61.4\%, 90.32\%) compared to non-enforcers estimate survival rate of 87.95\% (95\% CI: 83.09\%, 93.10\%).

@@

```{r estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for EP for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5."}
coxphmod_EP <- coxph(Surv ~ EP + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

@@
